---
title: "Bayesian Modelling Assessed homework"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

We install some packages and do clean up the data a little as well as split the dataset in two based on where the reading was taken.

```{r warning=FALSE, message=FALSE}
# we set the seed and install some packages

invisible(set.seed(999))
invisible(library(rjags, quietly = TRUE, warn.conflicts = FALSE))

# We load our data from the source
data <- read.table(file = "C:/Users/User/Documents/programming/Stats R work/datasets/DataSheet6/radon.txt", sep = '') 
# We convert our data to a data frame
data <- as.data.frame(data)

# we do some preprocessing on the data
county.list <- data[3]
data <- data[,c(1,2,4,5)]
head(data)
#we split the data based on if the house had a basement or not.
data_basement <- data[(data['floor'] == 1),]
data_floor <- data[(data['floor'] == 0),]
```



We are now ready to start modelling our data.


```{r results='hide'}

# We produce an intial values for the basement/first floor coefficients.
# Which by our model is the difference between the average of the basement and no basement readings 

beta <- NULL
for(i in 1:length(unique(data[3]))){
  beta0 <-c(beta, mean(data_basement[data_basement[,3] == i,1]) - mean(data_floor[data_floor[,3] == i,1])) 
}
beta0

# We define our model as shown on the accompanied sheet, and use non-informative priors

cat('
    model{
      for(i in 1:n){
        x[i] ~ dnorm(theta[c[i]] + beta * B[i], omega0^(-2))
      }
      for( j in 1:m){
        theta[j] ~ dnorm(alpha*uu[j], omega0.th^(-2))
      }
      alpha ~ dnorm(0, 0.001)
      beta ~ dnorm(beta0, 0.001)
      omega0 ~ dunif(0, 100)
      omega0.th ~ dunif(0, 100)
    }', file = 'radon.bug')

# we assign the data to the deterministic elements of the model
radon_data <- list(n = nrow(data), x = c(data[,1]), c = data[,3], B = data[,2], uu = unique(data[,4]), m = length(unique(data[,3])), beta0 = beta0 )

# we assign inital values to the stochastic elements of the model
radon_init <- list(alpha = 1, beta = beta0, omega0.th = 1, omega0 = 1, theta = rnorm(unique(data[,4])))

# We use jags to create the model
suppressMessages(log.radon <-jags.model('radon.bug', data = radon_data, init = radon_init))

```

Now that we have our model ready we can perform the monte carlo simulation. We first use 10000 iterations as a burn-out period for the model. Then we form a sample from the next 20000 iterations. 


```{r results = 'hide'}
# Our variables of interest are
log.radon.vars <- c("alpha", "beta", "omega0", "omega0.th", "theta")


# burn-out period
update(log.radon, n.iter = 10000, silent.jags = TRUE)

# We draw a sample after 20000 iterations of the model
sample <- coda.samples(log.radon, log.radon.vars, n.iter = 20000, silent.jags = TRUE)
```
Let's plot some graphs to check the burn-out period.
```{r}

# we set the plotting parameters to see if our burn-out period was a good choice.
par(mfrow = c(3,2))
pp <- NULL
for(i in 1:6){
  traceplot(sample[,i+5])
  pp <- c(pp, sample[,1+5])
}
# and the autocorrelations
autocorr.plot(pp, lag.max = 50)

```



Our selection of variables from the sample look good, the traceplot closely resembles white noise and the acf show a quick drop-off. This suggests that our choice of burn-out period was sufficient.
  We'll now look at the questions from the question sheet.

```{r}
par(mfrow = c(1,1))

average_log.radon <- NULL
observation_number <- NULL
for(i in 1:length(unique(data[,3]))){
  j <- data_basement[,3] == i
  # average with basement log.radon reading by county
  average_log.radon <- c(average_log.radon, mean(data_basement[j,1]))
  # number of with basement observations by county
  observation_number <- c(observation_number, nrow(data_basement[j,]))
}

plot(x = observation_number, y = average_log.radon, type = 'p', xlab = 'number of observations', main = 'Naive median log radon against number of observations')
minnesota_radon <- median(data_basement[,1])
abline(h = minnesota_radon)
```

```{r}
# we find the average for the values of alpha and beta from our MCMC simulation
est.alpha <- mean(sample[,1][[1]])
est.beta <- mean(sample[,2][[1]])
# We use the recorded uranium levels and our estimated value to estimate theta
est.theta <- est.alpha * unique(data[,4])
# We use our estimated parameters to estimate a mean for log.radon for each county
est.log.radon <- est.theta + est.beta
# and a estimated mean for the state
est.minnesota_radon <- median(est.log.radon)

plot(y = est.log.radon, observation_number, xlab = 'number of observations', main = 'MCMC median log radon against the number of observations' )
abline(h = est.minnesota_radon)

print(paste("We make a prediction of ", est.minnesota_radon, " for the log.radon in Minnesota."))
```
 
 
 We will now estimate the log radon in cook, a county with no observations. We use the model we defined and the parameters we estimated to find the readings in both the basement and the first floor.

```{r}
# cook uranium reading
cook.uu <- -0.504995982525783
# estimated readings
est.cook_firstfloor <- est.alpha*cook.uu 
est.cook_basement <- est.cook_firstfloor + est.beta
print(paste("The estimated log radon for a house in Cook with a basement is", est.cook_basement, ". The estimated log radon for a house in Cook with no basement is", est.cook_firstfloor, "."))
```
